Reproducible research in archaeology using R & rrtools

### 2018-03-10 analysis/paper/paper.Rmd
first try at compressing the text a bit.

OUr project comes from the group who took part in the 2017 Summer School on Reproducible Research in Landscape Archaeology at the Freie Universität Berlin (17-21 July), funded and jointly organized by Excellence Cluster 264 Topoi, the Collaborative Research Center 1266, and ISAAKiel.

@@

Our motivation for this topic comes from a profound change that we have been observing in how researchers analyse their data.

@@

## Slide: Current uses of R [signpost slide only]

@@

min_y <- 2007
max_y <- 2025
ggplot(cran_cites_all_areas_top_journals_by_year,
       aes(PY,
           prop,
           colour = SO)) +
  geom_line(size = 2) +
  geom_text_repel(
    data = subset(cran_cites_all_areas_top_journals_by_year, 
                  PY == max(PY)),
    aes(label = str_wrap(SO, 40)),
    size = 4,
    nudge_x = 0,
    hjust = 0,
    segment.color = NA,
    direction = "y"
  ) +
  scale_x_continuous(breaks = min_y:max_y,
                     labels = c(min_y:2017, rep("", length(2018:max_y))),
                     limits = c(min_y, max_y)) +
  scale_y_continuous(labels = scales::percent) +
  xlab("Publication year") +
  ylab("Percentage of articles in that journal") +
  theme_minimal() +
  theme(legend.position="none",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  ggtitle(str_glue('Percentage of articles per year citing R in top 10 WoS journals ({prettyNum(how_many_articles_in_top_journals_for_cran_cites,",")} articles)'),
          subtitle = "Data from apps.webofknowledge.com,\nusing in SCI-EXPANDED & SSCI for 2007-2017")

@@

This plot shows how data analysis is changing in the sciences more broadly towards the use of the R programming language. We searched the Web of Science for citations to the R program over the last ten years, and filtered the results to the top ten journals with the highest number of citations to this software. The purpose of this filtering step was to find areas of research were R is generally known, so we can see how it's use is changing over time. Note that many articles that use R do not cite the R program, so these numbers are underestimates of the true use of R in science.

We got `r how_many_articles_in_top_journals_for_cran_cites` articles in our sample, and we can make two major observations.

First, the areas of science that have seen big growth in the use of R are ecology, evolution and related biological sciences. This is especially significant for archaeologists, as ecology and biology are subjects from which they quite often adapt methods for analysis, as for example in point-pattern-analysis and statistical modelling. 
We also see increases in multi-disciplinary journals such as PeerJ and PLOS One.

**Second**, the scale of the increases is quite impressive. We see many journals in our sample starting from less than 1% of articles citing R and now have around 20% of articles citing R.

@@

Here are the results for archaeological articles. We found `r archy_journals_total_number_of_citing_articles` articles that cite R over the last ten years. Note that this is certinaly not complete, since these are just for the journals tracked by the Web of Science database, and we know that archaeologists publish is a great variaty of journals.

@@

The **first** detail we see here is that overall, the proportions of articles citing R in archaeology are much smaller than what we see in the ecology journals on the previous slide. The absolute numbers of archaeology articles here are are also very low. Only the _Journal of Archaeological Science_ has more than 10 articles in our sample. We have yet to adopt programming for data analysis in the same way that ecologists have.

The **second** detail is that we do not see a significant increase over time, unlike the ecology journals that steeply increase over time. Both archaeology and the other sciences we saw on the previous slide show an increase in citing R at around 2013-2014, we do not yet see much of an upward trend in the archaeology journals. Indeed, the first citations of R in several archaeology journals only occured in the last two years. The plot on the upper right of this slide shows a statistical test of change over time for articles in JAS. With a low r-squared value and high p-value for the linear model, we conclude that there is no evidence for an increase in citations over time in this journal.

@@

## Slide: The unrealised potential of R for reproducible research [signpost slide only]

@@

We would now like to focus on a small sub-set of these papers using R, because they are doing something very interesting that we think all archaeologists doing data analysis also should be doing. This subset of papers are demonstrating the potential of R for open and reproducible research.

In this figure we show a simple example of what we see in this small-subset of papers. On the left we see a plain-text document written in R Markdown. This means that is has lines of code in amongst lines of normal text. We can format the text with simple additions, such as the underscore to indicate italics and the pound symbol to indicate a heading. These simple additions are Markdown, and we can also use Markdown for citations and cross-references, though we don't show them in our example here. This document format allows us to mix R code and text in the same document. Here we see a short snippet of R code in the text (nrow(cars)) as well as a longer block of code below.

@@

This is how R helps to make research reproducible - we have found that some archaeologists are writing their entire journal articles useing R Markdown, and they share their plain-text document along with their publication. This means that the reader can see all of the details of the data analysis, and all of the small decisions made along the way. Many of these decisions are too tedious to describe in the methods section of an article, but they can be important for evaluating the robustness of an anlaysis. These details are also important to know if we want to extend a published analysis onto a new set of data.

@@

Here is a real-world example R Markdown in use, from an archaeology article published earlier this year.

@@

In our sample of archaeology articles, we identified `r number_of_reproducible_articles` articles that use R to make their work reproducible. These articles include R script files and data files with the articles, either in the supplementary materials or in a trustworthy data repository. This plot shows that although the number is small, it is increasing, suggesting the beginning of a new approach to how archaeologists share their research.

These `r number_of_reproducible_articles` articles in the subset of articles that use R to make their work reproducible are distinct from the other articles the cite R because most of the articles that cite R lack several features. These include:

@@

Articles that have these three characteristics are part of a new paradigm in scientific communication, and with these `r number_of_reproducible_articles` articles, we believe we have identified the beginning of this paradigm in archaeology. We think it will not be hard for archaeologists that already use R to make the transition to producing reproducible research by sharing their code and data. If this can be achieved, then we can realise some of the potential that R has for making archaeology more open and reproducible.

@@

@@

As we studied the papers in our sample, we looked for common approaches to making and organising the code and data files. We found that there were few common standards used by archaeologists in making their work reproducible. There was a great variety of information presented in the code and data files. Some papers include code that will generate every figure and numer in the text, while other papers only include code to produce one or two key figures in the paper.

@@

We have some more advanced functions that we will discuss more at another time.

>>>>

Our project comes from the group who took part in the 2017 Summer School on Reproducible Research in Landscape Archaeology at the Freie Universität Berlin (17-21 July), funded and jointly organized by Excellence Cluster 264 Topoi, the Collaborative Research Center 1266, and ISAAKiel.

@@

Our motivation for this topic arises from a profound change that we have been observing in how researchers analyse their data.

@@

## Slide: Current uses of R [signpost slide only]
So, let's start with the current uses of R.

@@

min_y <- 2007
max_y <- 2025
ggplot(cran_cites_all_areas_top_journals_by_year,
       aes(PY,
           prop,
           colour = SO)) +
  geom_line(size = 2) +
  geom_text_repel(
    data = subset(cran_cites_all_areas_top_journals_by_year, 
                  PY == max(PY)),
    aes(label = str_wrap(SO, 40)),
    size = 4,
    nudge_x = 0,
    hjust = 0,
    segment.color = NA,
    direction = "y"
  ) +
   annotate("text", x=2019, y=-0.01, label= "n = 50380") +
   scale_x_continuous(breaks = min_y:max_y,
                     labels = c(min_y:2017, rep("", length(2018:max_y))),
                     limits = c(min_y, max_y)) +
  scale_y_continuous(labels = scales::percent) +
  xlab("Publication year") +
  ylab("Percentage of articles in that journal") +
  theme_minimal() +
  theme(legend.position="none",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  ggtitle(str_glue('Percentage of articles per year citing R in top 10 WoS journals ({prettyNum(how_many_articles_in_top_journals_for_cran_cites,",")} articles)'),
          subtitle = "Data from apps.webofknowledge.com,\nusing in SCI-EXPANDED & SSCI for 2007-2017")

@@

This plot shows how data analysis is changing in the sciences more broadly towards the use of the R programming language. We searched the Web of Science for citations to the R program over the last ten years, and filtered the results to the top ten journals with the highest number of citations to this software. The purpose of this filtering step was to find areas of research were R is generally known, so we can see how it's use is changing over time. Note that many articles that use R do not cite the R program, so these numbers underestimate the true use of R in science.

We can make two major observations:

**First**, the areas of science that have seen big growth in the use of R are ecology, evolution and related biological sciences. This is especially significant for archaeologists, as ecology and biology are subjects from which we quite often adapt methods for analysis, as for example in point-pattern-analysis and statistical modelling. 
We also see increases in multi-disciplinary journals such as PeerJ and PLOS One.

**Second**, the scale of the increases is quite impressive: Starting only 2012 some journals now have 20% of articles citing R.

@@

Here are the results for archaeological articles. We found `r archy_journals_total_number_of_citing_articles` articles that cite R over the last ten years. Note that this is certainly not complete, since these are just for the journals tracked by the Web of Science database, and we know that archaeologists publish is a great variety of journals.

@@

The **first** detail we see here is that overall, the proportions and the absolute number of articles citing R in archaeology are much smaller than what we see in the ecology journals on the previous slide. Only the _Journal of Archaeological Science_ has more than 10 articles in our sample. It's obvious, we have yet to adopt programming for data analysis in the same way that ecologists have.

The **second** detail is that we do not see a significant increase over time. Both archaeology and the other sciences we saw on the previous slide show an increase in citing R at around 2013-2014, but we do not yet see much of an upward trend in the archaeology journals in contrast to e.g. ecology journals. Indeed, the first citations of R in several archaeology journals only occured in the last two years. The plot on the upper right of this slide shows a statistical test of change over time for articles in JAS. With a low r-squared value and high p-value for the linear model, we conclude that there is no evidence for an increase in citations over time in this journal.

@@

## Slide: The unrealised potential of R for reproducible research [signpost slide only]
After this overview how R is being used at the moment, we want to focus on the possibilities of making research reproducible using R.

@@

We would now like to focus on those archaeological papers using R, that are demonstrating the potential of R for open and reproducible research. These papers are using R Markdown for their research, which is shown in this figure.

On the left we see a plain-text document written in R Markdown. This means that is has lines of code in amongst lines of normal text. We can format the text with simple additions, such as the underscore to indicate italics and the hash symbol to indicate a heading. These simple additions are Markdown, a very easy markup language with plain text formatting syntax. It is designed so that it can be converted to HTML and many other formats. **R Markdown** allows us to mix R code and text in the same document. Here we see a short snippet of R code right inside the text (nrow(cars)) as well as a longer block of code below.

@@

We have found that some archaeologists are writing their entire journal articles useing R Markdown, and they share their plain-text R Markdown document along with their publication. This means that the reader can see all of the details of the data analysis, and all of the small decisions made along the way. This way the analysis is easily reproducible for others.

@@

Here is a real-world example R Markdown in use, from an archaeology article published in 2017.

@@

In our sample of R citing archaeology articles, we identified `r number_of_reproducible_articles` articles that include R script- and data files, either in the supplementary materials or in a trustworthy data repository. This plot shows that although this number is small, it is increasing, suggesting the beginning of a new approach to how archaeologists share their research.

The way in which these `r number_of_reproducible_articles` articles make their work reproducible is by:

@@

Articles that have these three characteristics are part of a new paradigm in scientific communication, and with these `r number_of_reproducible_articles` articles, we believe we may have identified the beginning of this behaviour in archaeology. We think it will not be hard for archaeologists that already use R to make the transition to producing reproducible research by sharing their code and data. If this can be achieved, then we can realise some of the potential that R has for making archaeology more open and reproducible.

@@

So, we developed a package for R, which assists researchers in taking this step.

@@

As we studied the papers in our sample, we looked for common approaches to making and organising the code and data files. We found that there were few common standards used by archaeologists in making their work reproducible. There was a great variety of information presented in the code and data files. Some papers include code that will generate every figure and number in the text, while other papers only include code to produce one or two key figures in the paper.

@@

Included in the package are some more advanced functions that we will discuss more at another time.
